---
title: 'Fuel Efficiency and Transmission Type: A Regression Analysis'
author: "Christian Salafia"
date: "2025-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(broom)
library(ggplot2)
library(performance)
library(car)
library(ggpubr)
```

# Executive Summary

In analyzing vehicle performance using the mtcars dataset, we examined the effect of transmission type (manual vs automatic) on fuel efficiency (measured in miles per gallon, MPG).

- Manual transmissions are associated with significantly higher MPG

- After adjusting for weight (wt) and horsepower (hp), manual cars average ~2.9 MPG more

- The difference is statistically significant (p = 0.037) with a 95% CI of [0.19, 5.69]

- Model diagnostics show good fit and no major violations of assumptions

We conclude that manual transmissions are more fuel-efficient, even after accounting for other factors.

# Data Preparation
```{r}
data("mtcars")

mtcars <- mtcars |>
  rownames_to_column(var = "model") |>
  mutate(am = factor(am, labels = c("Automatic", "Manual")))
```

# Modeling Strategy

We fit two models:

- Model 1 (Simple): mpg ~ am
- Model 2 (Adjusted): mpg ~ am + wt + hp

This approach allows us to:

- Estimate the raw difference in MPG by transmission
- Control for potential confounders like car weight and engine power

# Model Results

## Simple Linear Regression

```{r}
model_simple <- lm(mpg ~ am, data = mtcars)
tidy(model_simple)
```

## Adjusted Linear Regression

```{r}
model_adjusted <- lm(mpg ~ am + wt + hp, data = mtcars)
tidy(model_adjusted)
```

## Confidence Interval & Model Metrics

```{r}
confint(model_adjusted)["amManual", ]

glance(model_adjusted)
```

- The coefficient for amManual is statistically significant.
- 95% Confidence Interval does not cross 0 → strong evidence of effect.
- Adjusted R² ≈ 0.84 → excellent model fit

# Diagnostics

## Residual Plots

```{r residual-plot, fig.cap= "Residuals vs Fitted Values", fig.width=7, fig.height=5}
par(mfrow = c(2, 2))
plot(model_adjusted)
par(mfrow = c(1, 1))
```

## Assumptions Check
```{r check-model-bigger, fig.width=10, fig.height=8}
check_model(model_adjusted)
```

- Residuals are homoscedastic (constant variance).
- No major outliers or leverage issues.
- Q-Q plot suggests approximate normality

# Conclusions

Q1: Is a manual transmission better for MPG?

- Yes. Manual cars are significantly more fuel-efficient than automatics.

Q2: By how much?

- 2.94 MPG more on average, with a 95% confidence interval of [0.19, 5.69], even after adjusting for vehicle weight and horsepower.

Q3: How confident are we?

- p-value: 0.037 → statistically significant
- Model Fit: Adjusted R² = 0.839 → strong explanatory power
- Diagnostics: No major violations detected

# Appendix: Visualizations

## MPG by Transmission Type

```{r mpg-transmission-plot, fig.cap= "MPG by Transmission Type", fig.width=7, fig.height=5}
ggplot(mtcars, aes(x = am, y = mpg, fill = am)) +
  geom_boxplot(width = 0.5, alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, color = "black", alpha = 0.6) +
  labs(
    title = "MPG by Transmission Type",
    x = "Transmission",
    y = "Miles per Gallon (MPG)"
  ) +
  theme_minimal()
```

## Adjusted MPG with Weight & Horsepower Only

```{r adjusted-mpg-plot, fig.cap= "Adjusted MPG with Weight & Horsepower Only", fig.width=7, fig.height=5}
mtcars_aug <- augment(model_adjusted)

ggplot(mtcars_aug, aes(x = am, y = mpg)) +
  geom_boxplot(aes(fill = am), alpha = 0.4, outlier.shape = NA) +
  geom_jitter(aes(size = wt, color = hp), width = 0.1, alpha = 0.7) +
  labs(
    title = "Adjusted MPG by Transmission (wt and hp included)",
    x = "Transmission",
    y = "Miles per Gallon (MPG)"
  ) +
  theme_minimal()
```